---
title: "Hospital EDA"
output: html_document
date: "2025-02-04"
---

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(corrplot)
library(DataExplorer)

# Load the dataset
file_path <- "HospitalDurations.csv"
df <- read.csv(file_path, stringsAsFactors = FALSE)
```


```{r}
head(df)
```


```{r}
# Summary statistics
summary(df)
```


```{r}
# Check for missing values
colSums(is.na(df))
```

```{r}
# Histogram for each numeric variable
plot_histogram(df)
```


```{r}
# Boxplots for identifying outliers
plot_boxplot(df, by = "Region")  # Change "Region" as needed for categorical grouping
```


```{r}
library(corrplot)

# Correlation matrix for numeric variables
numeric_df <- df %>% select_if(is.numeric)
corr_matrix <- cor(numeric_df, use = "complete.obs")

# Plot the correlation matrix with correlation coefficients
corrplot(corr_matrix, 
         method = "color",    # Use colored tiles
         type = "upper",      # Show only the upper triangle
         addCoef.col = "black", # Add correlation coefficients in black
         tl.col = "black",    # Text label color
         tl.srt = 45,         # Rotate text labels
         number.cex = 0.8)    # Adjust size of the numbers
```


```{r}
# Pair plot for relationships
pairs(numeric_df)
```
```{r}
library(GGally)
# Select relevant columns: "Inf.Risk" and other numeric variables
columns_of_interest <- c("Inf.Risk", "Lgth.of.Sty", "Age", "R.Cul.Rat", "R.CX.ray.Rat", "N.Beds", "Avg.Pat", "Avg.Nur")
filtered_df <- numeric_df[, columns_of_interest]

# Pair plot focusing on "Inf.Risk"
ggpairs(filtered_df,
        columns = 1:ncol(filtered_df), # All columns in filtered_df
        title = "Pair Plot Relative to Infection Risk",
        lower = list(continuous = wrap("smooth", alpha = 0.3)),
        upper = list(continuous = wrap("cor", size = 4)),
        diag = list(continuous = wrap("densityDiag")))

# Save the plot to a file (optional)
ggsave("infection_risk_pair_plot.png", width = 12, height = 8)
```


```{r}
# Density plots for numerical features
plot_density(df)
```

EDA Results:
Log Transform:
    Lgth.of.Sty (if variance stabilization is needed).
    Inf.Risk (to normalize its distribution).
    N.Beds (if clustering or skewness affects the model).

Check Multicollinearity:
    With Avg.Pat and Avg.Nur being highly correlated, one of them might be excluded from the model.
```{r}
# Load necessary libraries
if (!require(car)) install.packages("car")
if (!require(MASS)) install.packages("MASS")
if (!require(DataExplorer)) install.packages("DataExplorer")
library(car)
library(MASS)
library(dplyr)
library(DataExplorer)

# Create new columns for log-transformed variables
df <- df %>%
  mutate(
    Lgth.of.Sty = log(Lgth.of.Sty),    # Log-transform length of stay
    N.Beds = log(N.Beds + 1),          # Log-transform number of beds
    Avg.Pat = log(Avg.Pat + 1),        # Log-transform number of patients
    Avg.Nur = log(Avg.Nur + 1)         # Log-transform number of nurses
  )

# Convert categorical variables to factors
df$Region <- as.factor(df$Region)
df$Med.Sc.Aff <- as.factor(df$Med.Sc.Aff)

# Plot density for all variables, including new log-transformed ones
plot_density(df)
```


```{r}
# Build the regression model
model <- lm(
  Lgth.of.Sty ~ Inf.Risk + Age + N.Beds + Avg.Nur + Avg.Pat +
    R.Cul.Rat + R.CX.ray.Rat + Region + Med.Sc.Aff,
  data = df
)

# Summarize the model
summary(model)
```

```{r}
# Check multicollinearity using VIF
vif(model)
```


```{r}
# Diagnostics: Plot residuals to check assumptions
par(mfrow = c(2, 2))
plot(model)
```

```{r}
# Adjusted R-squared
cat("Adjusted R-squared:", summary(model)$adj.r.squared, "\n")
```

Remove Avg.Pat because of high colinearity:

```{r}
model <- lm(
  Lgth.of.Sty ~ Inf.Risk + Age + N.Beds + Avg.Nur +
    R.Cul.Rat + R.CX.ray.Rat + Region + Med.Sc.Aff,
  data = df
)
```

```{r}
summary(model)
```


```{r}
# Build the regression model
model <- lm(
  Lgth.of.Sty ~ Inf.Risk + Age + N.Beds + Avg.Nur + Avg.Pat +
    R.Cul.Rat + R.CX.ray.Rat + Region + Med.Sc.Aff,
  data = df
)

# Summarize the model
summary(model)
```


```{r}
# Adjusted R-squared
cat("Adjusted R-squared:", summary(model)$adj.r.squared, "\n")
```

```{r}
# Calculate Cook's Distance for each observation
cooks_d <- cooks.distance(model)

# Define a threshold for influential points (commonly 4 / n)
threshold <- 4 / nrow(df)

# Identify and remove highly influential observations
influential_points <- which(cooks_d > threshold)
cat("Number of influential points detected:", length(influential_points), "\n")

# Remove influential observations from the dataset
df_no_outliers <- df[-influential_points, ]

# Check the size of the dataset after removing outliers
cat("Original dataset size:", nrow(df), "rows\n")
cat("Dataset size after removing influential points:", nrow(df_no_outliers), "rows\n")

# Optional: Rebuild the model with the cleaned dataset
final_model <- lm(
  Lgth.of.Sty ~ Inf.Risk + Age + N.Beds + Avg.Nur +
    R.Cul.Rat + R.CX.ray.Rat + Region + Med.Sc.Aff,
  data = df_no_outliers
)

summary(final_model)
  #mae

# Make predictions on df
predicted <- predict(final_model, newdata = df)

# Calculate MAE properly
mae_value <- mean(abs(predicted - df$Lgth.of.Sty))  # Direct base R method

mae_value


# Diagnostics: Plot residuals to check assumptions
par(mfrow = c(2, 2))
plot(final_model)
```

```{r}
# Adjusted R-squared
cat("Adjusted R-squared:", summary(model)$adj.r.squared, "\n")
```

Creating Model with Interaction Terms

```{r}



model_int <- lm(
  Lgth.of.Sty ~ Inf.Risk + Age + N.Beds + Avg.Nur +
    R.Cul.Rat + R.CX.ray.Rat + Region + Med.Sc.Aff + Region * Age + Avg.Pat * N.Beds,
  data = df_no_outliers
)

summary(model_int)

```
```{r}
vif(model_int, type = 'predictor')
```

VIF's, checking assumptions



Evaluating performance of Model with Interaction Terms

```{r}
summary(model_int)


# Make predictions on df
predicted <- predict(model_int, newdata = df)

# Calculate MAE properly
mae_value <- mean(abs(predicted - df$Lgth.of.Sty))  # Direct base R method

mae_value
```

##############################################################
KNN
##############################################################

```{r}
# Split the data into training and testing sets
set.seed(123) # For reproducibility
train_index <- createDataPartition(df_no_outliers$Lgth.of.Sty, p = 0.8, list = FALSE)
train_data <- df_no_outliers[train_index, ]
test_data <- df_no_outliers[-train_index, ]
```


```{r}
# Standardize the numerical variables
preprocess <- preProcess(train_data[, c("Inf.Risk", "Age", "N.Beds", "Avg.Nur", "Avg.Pat", "R.Cul.Rat", "R.CX.ray.Rat")],
                         method = c("center", "scale"))
train_data_scaled <- train_data
test_data_scaled <- test_data

train_data_scaled[, c("Inf.Risk", "Age", "N.Beds", "Avg.Nur", "Avg.Pat", "R.Cul.Rat", "R.CX.ray.Rat")] <- predict(preprocess, train_data[, c("Inf.Risk", "Age", "N.Beds", "Avg.Nur", "Avg.Pat", "R.Cul.Rat", "R.CX.ray.Rat")])
test_data_scaled[, c("Inf.Risk", "Age", "N.Beds", "Avg.Nur", "Avg.Pat", "R.Cul.Rat", "R.CX.ray.Rat")] <- predict(preprocess, test_data[, c("Inf.Risk", "Age", "N.Beds", "Avg.Nur", "Avg.Pat", "R.Cul.Rat", "R.CX.ray.Rat")])
```


```{r}
# Train the KNN model with cross-validation
set.seed(123)
knn_model <- train(
  Lgth.of.Sty ~ Inf.Risk + Age + N.Beds + Avg.Nur + Avg.Pat + R.Cul.Rat + R.CX.ray.Rat + Region + Med.Sc.Aff,
  data = train_data_scaled,
  method = "knn",
  tuneGrid = data.frame(k = seq(1, min(nrow(train_data_scaled), 10), by = 2)), # Test k = 1, 3, 5, ..., up to 10 or dataset size
  trControl = trainControl(method = "cv", number = 5) # 5-fold cross-validation
)

# View the best k and model summary
print(knn_model)
```


```{r}
# Make predictions on the test set
knn_predictions <- predict(knn_model, newdata = test_data_scaled)

# Evaluate model performance
model_performance <- postResample(pred = knn_predictions, obs = test_data_scaled$Lgth.of.Sty)
cat("\nModel Performance on Test Set:\n")
print(model_performance)

# Plot the predicted vs actual values
ggplot(data.frame(Predicted = knn_predictions, Actual = test_data_scaled$Lgth.of.Sty), aes(x = Actual, y = Predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "KNN Predicted vs Actual Values", x = "Actual", y = "Predicted") +
  theme_minimal()


```







