---
title: "Project 1 EDA"
author: "Robert Carstens"
date: "2025-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Looking at dataset

```{r cars}
head(Hospital)
View(Hospital)

library(GGally)
ggpairs(Hospital)
```



```{r cars}
cor_matrix <- cor(Hospital[sapply(Hospital, is.numeric)], use="complete.obs")
cor_matrix["Lgth.of.Sty", ]


#after some initial EDA one thing is clear, Region should not be numeric
`````{r cars}
cor_matrix <- cor(Hospital[sapply(Hospital, is.numeric)], use="complete.obs")
cor_matrix["Lgth.of.Sty", ]


#after some initial EDA one thing is clear, Region should not be numeric
````


Steps to fix data
We need to make region a factor and not a numeric
```{r cars}
Hospital$Region <- as.factor(Hospital$Region)

library(ggplot2)
ggplot(Hospital, aes(x = Region, y = Lgth.of.Sty)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Length of Stay by Region") +
  theme_minimal()

```

Even though conventional wisdom would say region should most likely be a factor and not a numeric. It looks like there is kind of a linear relationship


VIFS

```{r cars}

library(car)


model <- lm(Lgth.of.Sty ~ Age + Inf.Risk + R.Cul.Rat + R.CX.ray.Rat + 
             N.Beds + Med.Sc.Aff + Avg.Pat + Avg.Nur + Pct.Ser.Fac + Region, 
             data = Hospital)


improved_model <- lm(Lgth.of.Sty ~ Age + Inf.Risk + R.Cul.Rat + R.CX.ray.Rat + 
             N.Beds + Med.Sc.Aff   + Pct.Ser.Fac + Region, 
             data = Hospital)

vif(model)
vif(improved_model)
cor(Hospital$N.Beds, Hospital$Avg.Nur)





```
```{r cars}
library(GGally)

# Select only the variables in your improved_model
selected_vars <- c("Lgth.of.Sty", "Age", "Inf.Risk", "R.Cul.Rat", "R.CX.ray.Rat", 
                   "N.Beds", "Med.Sc.Aff", "Pct.Ser.Fac", "Region")

# Subset the dataset
Hospital_subset <- Hospital[selected_vars]

# Generate the pair plot
ggpairs(Hospital_subset)


```

```{r cars}
improved_model <- lm(Lgth.of.Sty ~ Age + Inf.Risk + R.Cul.Rat + R.CX.ray.Rat + 
             N.Beds + Med.Sc.Aff   + Pct.Ser.Fac + Region, 
             data = Hospital)
summary(improved_model)
```